---
title: "**Data**"
author: "Anna Beijl"
date: "17 September 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# First ideas

What type of data do I need?

-   3 types of groups, at least 2 different outgroups

<br>

# Data options

*The Nijmegen School Study (TNSS)*

-   Vooroordelen en positief/negatief contact obv etniciteit en gender

    -   O.a. thermometerschalen, percentages, ook contact ouders

-   'Hoeveel van je goede vrienden zijn x?' –\> is dit genoeg (compleet) netwerkdata?

    -   Matrix lijkt alleen over pesten, ingevuld door mentoren

    -   Pesten dan als behavioral DV nemen?

*The Arnhem School Study (TASS)*

-   Geen toegang tot codeboek online (<https://ssh.datastations.nl/dataset.xhtml?persistentId=doi:10.17026/DANS-ZC6-JHDK>) –\> via Jochem?

*FIS data*

-   1/2/3 generation migrants (but maybe too in-group-y?)

-   Intergroup attitudes based on religion and ethnicity

-   Other option: Perceived discrimination

    -   Also only based on migration background / ethnicity though

    -   Possibly extend logic of social influence effects to perceived discrimination?

    -   Some literature: <https://doi.org/10.1177/21568693241271289>, <https://doi.org/10.1037/pspi0000426>, <https://link.springer.com/article/10.1007/s10964-023-01746-1>

-   But: a LOT less literature + it would mean to drop the secondary group thing

*MRCS data*

-   Salience of social identity elements

-   Intergroup contact attitudes (broad!) \> "people of other social identities"

<br>

# Data descriptions

Should include:

-   Context and participants

-   Waves: how many, what years?

-   Directed / undirected adjacency matrix

-   DV and IVs

    -   The dependent variable in the stochastic actor oriented model (SOAM) is the network structure in wave two and three

        -   In mijn geval dus niet! Alleen een behavioral DV en netwerk is vaststaand

    -   Ego-level (monadic) covariates: constant/varying

    -   IV: attitudes of ego / students towards [secondary outgroup]

-   Interaction model? Ego's attitudes lead to ego's friendships, but this is conditional on friends' (from primary outgroup) opinions

    -   Two-step influence effects?

-   Control variables –\> structural network effects such as transitivity / triadic closure, reciprocity, inPop, outAct, isolateNet

-   Corresponding to which effects (i.e. technical names) in the RSiena model?

<br>

# Updated RQs

-   **Descriptive (macro / network level)** \> What is the diversity in students' social networks based on [secondary outgroup]?

-   **Explanatory (micro level)** \> To what extent are secondary out-group attitudes / friendships (diversity in network) explained by *social influence effects* (vs. secondary transfer effects of intergroup contact)?

    -   What is the primary vs secondary outgroup?

        -   Similar in stigma/marginalization?? (condition for STE)

    -   Is the outgroup fixed (e.g. always 'women' and 'ethnic minorities') or dynamic based on who ego is (e.g. for an ethnic minority woman, the outgroups would be 'men' and 'ethnic majorities')?

    -   Selection vs. influence no longer possible?

<br>

# Describing my data

## Loading / preparing data 

```{r}
rm(list = ls()) # clean workspace

# custom functions
fpackage.check <- function(packages) {
    lapply(packages, FUN = function(x) {
        if (!require(x, character.only = TRUE)) {
            install.packages(x, dependencies = TRUE)
            library(x, character.only = TRUE)
        }
    })
}

fsave <- function(x, file = NULL, location = "./data/") {
    ifelse(!dir.exists("data"), dir.create("data"), FALSE)
    ifelse(!dir.exists("data"), dir.create("data"), FALSE)
    if (is.null(file))
        file = deparse(substitute(x))
    datename <- substr(gsub("[:-]", "", Sys.time()), 1, 8)
    totalname <- paste(location, datename, file, ".rda", sep = "")
    save(x, file = totalname)  #need to fix if file is reloaded as input name, not as x. 
}

fload <- function(filename) {
    load(filename)
    get(ls()[ls() != "filename"])
}

fshowdf <- function(x, ...) {
    knitr::kable(x, digits = 2, "html", ...) %>%
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
        kableExtra::scroll_box(width = "100%", height = "300px")
}

packages = c("RSiena", "tidyverse", "stringdist", "stringi")

fpackage.check(packages)

library(haven)
tnss_net <- read_sav("./data/TNSS_wave_2010_mentors.sav")
tnss_pup10 <- read_sav("./data/TNSS_wave_2010_pupils.sav")
tnss_pup11 <- read_sav("./data/TNSS_wave_2011_pupils.sav")
# using 'foreign' I got warnings and didn't seem to load well. 'haven' works nicely

view(tnss_net)
view(tnss_pup10)
view(tnss_pup11) # all looks good

class(tnss_net) # tibble/data frame
class(tnss_pup10)
class(tnss_pup11)

#join datasets
# gaat automatisch goed gelukkig (obv IDClass en IDresp)
tnss <- left_join(tnss_net, tnss_pup10)
view(tnss)

#selecteer class 5 (of een andere leuke grote klas met weinig missings)
c5 <- tnss %>% 
  filter(IDClass == 5)

c9 <- tnss %>% 
  filter(IDClass == 9)

c3 <- tnss %>% 
  filter(IDClass == 3)

#aantal leerlingen in class 5
max(c5$order) # 28
max(c9$order) # 20
max(c3$order) # 27

#vis het nominatienetwerk er uit. Nu beetje handwerk. Kan vast slimmer. 
c5_net <- c5 %>%
  select(fr_1:fr_28)
c3_net <- c3 %>%
  select(fr_1:fr_28)
view(c5_net)
class(c5_net) # tibble/data frame
```

## Descriptive stats

```{r}
require(igraph)
g1 <- graph_from_adjacency_matrix(as.matrix(c5_net))
#kleur op basis van gender. staat in goede volgorde. 
V(g1)$color <- ifelse(c5$gender == 1, "red", "green")
plot(g1)

vcount(g1) # count vertices / actors
ecount(g1) # count edges / ties
is_bipartite(g1) # is the network unimodal or multi? # F
is_weighted(g1) # F

require(RsienaTwoStep)
ts_recip(g1) # 68
ts_outAct(g1) # 6724
ts_inAct(g1) # same number
triad_census(g1)
dyad_census(g1)
# $mut = number of pairs with mutual connections 
# $asym = number of pairs with non-mutual connections 
# $null = number of pairs with no connections between them 

gmat <- as_adjacency_matrix(g1, type = "both", sparse = FALSE)
isSymmetric(gmat) # F, dus directed!
```

## Visualization

```{r}
# transivitity
igraph::transitivity(g1, type = c("localundirected"), isolates = c("NaN", "zero"))

# betweenness
igraph::betweenness(g1, directed = T)


# Let’s make size proportional to betweenness score:
# changing V
V(g1)$size = betweenness(g1, normalized = T, directed = T) * 60 + 10  #after some trial and error
plot(g1, mode = "directed")

view(tnss_pup10)

# making it prettier
#GENDER
plot(g1 ,
     vertex.color = ifelse(c5$gender == 1, "hotpink", "olivedrab2"), # gold = nice color for the vertices
     vertex.size = betweenness(g1, normalized = T, directed = T) * 60 + 10,  # 20 we'll vertices a bit smaller
     vertex.frame.color = "gray",  # we'll put a gray frame around vertices
     vertex.label.color = "black",  # not that ugly blue color for the labels (names)
     vertex.label.family = "Helvetica", # not a fan of times new roman in figures
     vertex.label.cex = 0.5,  # make the label a bit smaller too
     vertex.label.dist = 0.8,  # we'll pull the labels a bit away from the vertices
     edge.curved = 0, # curved edges is always a nice touch
     edge.arrow.size = 0.4) # make arrow size (direction of edge) smaller

plot(g1 ,
     vertex.color = ifelse(c5$ethnicity == 1, "orange", "olivedrab2"), # gold = nice color for the vertices
     vertex.size = betweenness(g1, normalized = T, directed = T) * 60 + 10,  # 20 we'll vertices a bit smaller
     vertex.frame.color = "gray",  # we'll put a gray frame around vertices
     vertex.label.color = "black",  # not that ugly blue color for the labels (names)
     vertex.label.family = "Helvetica", # not a fan of times new roman in figures
     vertex.label.cex = 0.5,  # make the label a bit smaller too
     vertex.label.dist = 0.8,  # we'll pull the labels a bit away from the vertices
     edge.curved = 0, # curved edges is always a nice touch
     edge.arrow.size = 0.4) # make arrow size (direction of edge) smaller
```

What do we see:

-   A lot of clustering!

    -   It seems strongly based on gender

    -   Also (but less) based on ethnicity

-   A couple of students with really high levels of betweenness (15, 14, 4, 3) –\> connecting the different clusters; brokers / bridges

    -   One very central actor: 15 (male + Dutch origin)

    -   Not much variation among the other students

<br>

## Defining data + RSiena object

tbd later

```{r}
# diag(c5_net) <- 0
length(tnss) #173
length(c5_net)
# hoe verder, want ik heb maar 1 wave
# tnss_array <- array(data = c(wave1, wave2), dim = c(dim(wave1), 2))
# net <- sienaDependent(tnss_array)

# INDEPENDENT VARIABLE ----------------------------------------------------
# gender
# gender <- as.numeric(c5$gender == "female")
# gender <- coCovar # = constant covariate, kan ook dyad, time varying, etc

# samenvoegen in één data object voor Siena
# mydata <- sienaDataCreate(net, gender)
```

## Effects structure and initial description

tbd later

```{r}
# myeff <- getEffects(mydata)
# effectsDocumentation(myeff) # dit zijn alle mogelijke effecten die je kunt opnemen voor dit data object
# ifelse(!dir.exists("results"), dir.create("results"), FALSE)
# print01Report(mydata, modelname = "./results/tnss") # goed bekijken
```
