---
title: "Week1"
author: "Anna Beijl"
date: "2026-08-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workshop 31-08-26

## Summary

In the workshop we did the following:

-   Intro in the course and final assignment

-   Intro in some social network concepts / theory

-   Intro inRmd and setting up the lab journal and website

I started with the R tutorial in Appendix A of the SNASS textbook. I also started thinking about research questions, and started looking for literature.

## Current to do list

-   Think about RQs
-   Look for literature and pick two articles to present
-   Finish the tutorial, Appendices A and B
-   Finish the lab journal for this week
-   Read chapters in SNASS textbook

# Lab journal

## Setting up the lab journal

```{r}
# install.packages("rmarkdown", dependencies = TRUE) 
# install.packages("usethis") 
library("usethis")
use_git_config(user.name = "annabeijl", user.email = "anna.beijl@ru.nl") 
create_github_token() 
gitcreds::gitcreds_set() 
```

## Working through the R tutorial 

From the SNASS textbook, Appendix A

```{r}

```
