---
title: "**Lab journal week 1**"
author: "Anna Beijl"
date: "02 September 2026"
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workshop 31-08-26

## *Summary*

In the workshop we did the following:

-   Intro to the course and final assignment

-   Intro to some social network concepts / theory

-   Intro to Rmd / Github and setting up the lab journal and website

I started working through the R tutorial in Appendix A of the SNASS textbook. I also started thinking about research questions, and started looking for literature.

## *Current to do list*

-   Think about RQs
-   Look for literature and pick two articles to present
-   Finish Appendices A and B
-   Finish the lab journal for this week
-   Read chapters 1, 2, 3.3, 4 in SNASS textbook

# R and Github

## *Setting up the lab journal*

```{r warning=FALSE}
# install.packages("rmarkdown", dependencies = TRUE) 
# install.packages("usethis") 
library("usethis")
# use_git_config(user.name = "annabeijl", user.email = "anna.beijl@ru.nl") 
# create_github_token() 
# gitcreds::gitcreds_set()
```

## *Working through the R tutorial*

From the SNASS textbook, Appendix A. I saved the R code in a different script, to keep this one clean and legible.

# RQs and literature

## *Two articles*

1.  @zingora2026

    *Outgroup friendships and social influence in the development of adolescent attitudes toward secondary outgroups*

    1.  **Central question**: secondary transfer effects \> due to generalization / transfer of intergroup contact (classic idea), or actually social influence as friends shape each other's outgroup attitudes?
    2.  **Theory**
        -   Social influence is not just positive! If your friend from outgroup A is negative about outgroup B; you will become more negative too

        -   Adolescents adopt friends' attitudes because of e.g. frequent observation, to avoid social exclusion

        -   It is possible that previously documented STEs (including null-effects) are actually due to social influence!

        -   Hypotheses

            -   Classic intergroup contact hypos: number of outgroup A friends –\> number of outgroup B friends

            -   Social influence hypos: *attitudes* toward outgroup B of outgroup A friends –\> attitudes of ego toward outgroup B
    3.  **Data**: 3 studies, 2 empirical and 1 simulation. Longitudinal network data on adolescents (in NL *n* = 812, and DE *n* = 2,559)
        -   *"Whereas SAOM allows us to test which processes take place in adolescent social networks, the ABM simulation approach enables us to capture the implications of these simultaneous processes for the development of the wider network and the spread of outgroup attitudes (1) and to further advance theory by generating hypotheses for future research (2)"* –\> Micro-macro link

        -   The Arnhem School Study (TASS) and adolescents in Germany, 2013-2014

        -   Measures: intergroup attitudes *and* friendships
    4.  **Method**: stochastic actor-oriented models in RSiena
        -   Control for effects of ingroup friends on secondary outgroup attitudes + network dynamics

        -   Simulation: to test whether the social influence mechanisms vary for adolescents with negative, neutral, and positive attitudes when their primary outgroup friends hold negative, neutral, or positive attitudes
    5.  **Results**:
        -   Social influence mechanism supported \> 'attitudes on secondary outgroups become more similar to those of primary outgroup friends'

        -   Secondary transfer effect of intergroup contact *not supported* \> 'attitudes on secondary ougroups become more inclusive when having more outgroup friends'

            -   These findings hold for both ethnic majority and ethnic minority adolescents!

-   Social networks perspective:

    -   **Theoretical mechanisms**: very present, new angle in this field; focus on social influence and interdependence

    -   **Data**: complete network data? idk...

        -   "Adolescents (N=1,314,12–13 years old, 46% female) completed an online questionnaire in their school’s computer lab under a teacher’s supervision. Data were collected in 58 classrooms nested in 11 schools."

    -   **Method**: seems great

2.  @chávez2021

    *The role of perspective-taking and low social class prejudice on cross-ethnic friendship formation*

    1.  **Central question**: Are people with low prejudice on social class level + perspective-taking abilities, also more likely to have low prejudice on ethnicity level? And thus engage in cross-ethnic friendships?

        -   My interest: intersectionality social class x ethnicity / secondary transfer effects towards other out groups?

    2.  **Theory**: social networks perspective

        -   Micro-level explanatory mechanisms: propinquity (availability of options)? (ethnic) homophily? Individual attitudes! Perspective-taking + social class prejudice

        -   A multicultural context does not imply that positive inter-group contact will take place. Instead, we can expect that, by default, individuals will spend time and interact with peers from their own ingroup. Thus: focus on *perspective taking* and *prejudice* \> And shared context effects are not likely

        -   Intersectional?

            -   "Social class shapes and affects the life experiences regarding other social categories, such as ethnicity" \> "Members of minority groups in Chile (i.e., indigenous and immigrant students) are usually seen as belonging to a lower status" \> "the discriminatory ethnic experience could be mainly based on the socioeconomic exclusion of these marginalized groups"

    3.  **Data:** 242 students (M age = 12, all relatively low SES) from ﬁve multicultural classrooms in Chile, longitudinal data on 3 time points, from 2017

    4.  **Method**: longitudinal multiplex social network models implemented in RSiena

        -   Controlled for structural network effects: reciprocity, transitivity, ego/alter/similarity effects, indegree popularity, in-degree activity

    5.  **Results**: all hypotheses supported

        -   So: yes, homophily

        -   But also: more cross-ethnic friendship among students with high perspective taking and low social class prejudice

        -   "When social class prejudice is low, this allows students to break down the barrier of ethnicity and, consequently, to build friendships with low social class peers who, importantly, belong to a different ethnic group"

-   Social networks perspective

    -   **Hypotheses and theoretical mechanisms**: relationships

        -   But: individual attributes?

        -   Remains on the micro level; no descriptive macro level RQ

    -   **Data**: complete network / socionet, longitudinal -\> great

    -   **Method**: also seems fitting, don't know enough RSiena yet

## *First setup RQs*

-   **Descriptive (macro / network level)** \> *Diversity* in students' social networks based on [secondary outgroup]? What is the *distribution* of secondary out-group attitudes / friendships in the students' network?

-   **Explanatory (micro level)** \> To what extent are secondary out-group attitudes / friendships (diversity in network) explained by *social influence effects* (vs. secondary transfer effects of intergroup contact)?

    -   Macro X / starting conditions?

    -   Social influence: where in the GTF? Arrow instead of micro X or Y?

-   Who are the outgroups?

    -   From ethnic minority perspective?

    -   Completely different type of outgroup, not based on ethnicity?
