---
title: "**Lab Journal Week 4**"
author: "Anna Beijl"
date: "21 September 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workshop 21-09-2026

## *Summary*

In the workshop we did the following:

-   A few student presentations on RQ and data

-   Intro to network visualization

-   Intro to web scraping

-   Working with our own data, making first few plots

## *Current to-do list:*

-   Repeat everything from this (and previous) week(s)
-   Tidy up notes, lab journal, and website
-   SNASS chapter 8 and 9
-   Edit RQs
-   Prepare data for RSiena analysis
    -   Descriptive stats
    -   Visualization
    -   Think about modeling strategy
-   Read papers on stationary SAOM / figure out analysis

<br>

# Lab journal

## Network visualization

Based on the tutorial in @SNASS chapter 9

### Zachary’s karate club

#### *Loading data*

```{r}
# loading the data - Zachary's karate club
require(igraph)
g <- make_graph("Zachary")
plot(g)
# 'g' is een igraph object

# adjacency matrix van maken
gmat <- as_adjacency_matrix(g, type = "both", sparse = FALSE)
gmat
isSymmetric(gmat) # symmetrisch, dus undirected!
```

#### *Descriptive statistics*

```{r}
# size

# degree

# TRANSITVITY
# hoeveel gesloten driehoekjes van alle mogelijke driehoekjes?
# be aware that directed graphs are considered as undirected. but g is undirected.
igraph::transitivity(g, type = c("localundirected"), isolates = c("NaN", "zero"))

# BETWEENNESS
# hoe vaak zit ego op de kortste paden tussen 2 actoren
igraph::betweenness(g, directed = FALSE)
# actor 1, 3 en 34 springen er heel erg uit, maar op het plaatje zie je niet veel verschil tussen 33 en 34. hier kun je mee spelen in het plot; als je betweenness centraal wil stellen, wil je meer nadruk leggen op actoren 1 en 34. bijvoorbeeld: door andere kleuren/intensiteit, grootte of layout obv betweenness

igraph::dyad_census(g)

igraph::triad_census(g)
# I will use sna because it shows the names of the triads as well.
sna::triad.census(gmat)
unloadNamespace("sna")  #I will detach this package again, otherwise it will interfere with all kind of functions from igraph, and my students will hate me for that.

# GLOBAL TRANSITIVITY
igraph::transitivity(g, type = "global")
sna::gtrans(gmat)

triad_g <- data.frame(sna::triad.census(gmat))

transitivity_g <- (3 * triad_g$X300)/(triad_g$X201 + 3 * triad_g$X300)
transitivity_g
```

We zien nu dat er verschillen zijn in lokale transitiviteit en in betweenness, en er zijn 2 nodes die eruit springen (= met hoge betweenness, lage local transitivity). Maar die twee nodes zijn niet met elkaar verbonden; interessant!

#### *Network visualization*

```{r}
# Let’s make size proportional to betweenness score:
# changing V
V(g)$size = betweenness(g, normalized = T, directed = FALSE) * 60 + 10  #after some trial and error
plot(g, mode = "undirected")

# we willen actor 1 en 34 toch nog wat verder uit elkaar. dit doen we obv een algoritme
set.seed(2345)
l <- layout_with_mds(g)  #https://igraph.org/r/doc/layout_with_mds.html
plot(g, layout = l)

l  #let us take a look at the coordinates
l[1, 1] <- 4
l[34, 1] <- -3.5
plot(g, layout = l)

# notitie toevoegen om te laten zien dat we hebben 'gemanipuleerd':
plot(g, layout = l, margin = c(0, 0, 0, 0))
legend(x = -2, y = -1.5, c("Note: the position of nodes 1 and 34 have been set by Jochem Tolsma \n for visualisation purposes only and do not reflect network properties"),bty = "n", cex = 0.8)
```

### Co-author networks

#### *Loading data*

```{r}
# load the necessary datasets
load("./data/soc_df.RData")
load("./data/soc_collabs1.RData")
load("./data/soc_collabs2.RData")

# get unique collaborators of soc staff first
soc_collabs_unique <- unique(soc_collabs[, 4])  # so 229 unique collaborators for RU staff?
soc_collabs_unique <- c(soc_collabs_unique, soc_df$gs_id)  # add the soc staff themselves.
soc_collabs_unique <- data.frame(soc_collabs_unique)
soc_collabs_unique$v1 <- 1  # convenient to select on after the merge
soc_collabs_unique <- unique(soc_collabs_unique)

# so this is a very important step, we join the unique soc collaborators to the collaborators of collaborators
require(tidyverse)
onefivedegree <- left_join(collabs_1deep, soc_collabs_unique, by = c(coauth_id = "soc_collabs_unique"))

# Then, we drop those that are not among sociology collaborators and who don't lsit coauthors
# regardless
onefivedegree <- onefivedegree[!is.na(onefivedegree$v1), ]
onefivedegree <- onefivedegree[!is.na(onefivedegree$coauth), ]

# we pick those columns and have an edgelist of soc collaborators and whether they collaborate with
# those same collaborators the 1.5 degree network
onefivedegree <- onefivedegree[, c("name", "coauth")]
names(onefivedegree) <- c("from", "to")

# we get soc collaborators and add those to the data above and end up with a nice edgelist!
socc <- soc_collabs[!is.na(soc_collabs$coauth), ]
socc <- socc[, c("name", "coauth")]
names(socc) <- c("from", "to")
onefivedegree <- rbind(socc, onefivedegree)
save(onefivedegree, file = "./data/soc_onefivedegree.RData")
```

#### *Network visualization*

```{r}
onefivedegree <- as.matrix(onefivedegree)  # matrix because igraph wants that
library(igraph)  # Notice that we call igraph here, in anticipation of the viz tutorial.
net1 <- graph_from_edgelist(onefivedegree, directed = TRUE)  # Notice the igraph function here!

plot(net1)  # pretty ugly!
```

#### *Making it prettier*

```{r}
plot(net1 ,
     vertex.color = "gold", # nice color for the vertices
     vertex.size = 4,  # we'll vertices a bit smaller
     vertex.frame.color = "gray",  # we'll put a gray frame around vertices
     vertex.label.color = "black",  # not that ugly blue color for the labels (names)
     vertex.label.family = "Helvetica", # not a fan of times new roman in figures
     vertex.label.cex = 0.4,  # make the label a bit smaller too
     vertex.label.dist = 0.5,  # we'll pull the labels a bit away from the vertices
     edge.curved = 0.2, # curved edges is always a nice touch
     edge.arrow.size = 0.1) # make arrow size (direction of edge) smaller
```

#### *Extracting node attributes*

Selecting Radboud sociology nodes and give them a different color.

```{r}
in_network <- data.frame(as_ids(V(net1)))
names(in_network)[1] <- "name"

#soc_df$total_cites <- soc_df$total_cites.x
ru_nodes <- soc_df[, c("name", "total_cites")]
in_network <- left_join(in_network, ru_nodes, by = c("name" = "name"))
in_network$vcol <- ifelse(is.na(in_network$total_cites), "#E69F00", "#56B4E9")

plot(net1,
     vertex.color = in_network$vcol, #THIS WAS WHAT WE DID THE LAST CODEBLOCK FOR!
     vertex.size = 4,  # we'll make them a bit smaller
     vertex.frame.color = "gray",  # we'll put a frame around it
     vertex.label.color = "black",  # not that ugly blue color for the names
     vertex.label.family = "Helvetica", # not a fan of times in figures
     vertex.label.cex = 0.4,  # a bit smaller too
     vertex.label.dist = 0.5,  # we'll pull the labels a bit away from the vertices
     edge.curved = 0.2,  # curved edges is always a nice tough
     edge.arrow.size = 0.1) # arrow size smaller
```

Show only the labels of RU sociology staff

```{r}
plot(net1,
     vertex.color = in_network$vcol, 
     #NOTICE THESE CONDITIONAL STATEMENTS BELOW
     vertex.label = ifelse(!is.na(in_network$total_cites), in_network$name, NA),
     vertex.size = 4,  # we'll make them a bit smaller
     vertex.frame.color = "gray",  # we'll put a frame around it
     vertex.label.color = "black",  # not that ugly blue color for the names
     vertex.label.family = "Helvetica", # not a fan of times in figures
     vertex.label.cex = 0.65,  # a bit smaller too
     vertex.label.dist = 0.5,  # we'll pull the labels a bit away from the vertices
     edge.curved = 0.2, # curved edges is always a nice tough
     edge.arrow.size = 0.1) # arrow size smaller
```

Vertex size adjusted for citation

```{r}
plot(net1,
     vertex.color = in_network$vcol, 
     vertex.label = ifelse(!is.na(in_network$total_cites), in_network$name, NA),
     # SAME HERE, TRY TO SMOOTH THE TOTAL_CITES A BIT WITH LOGSCALE
     vertex.size = ifelse(!is.na(in_network$total_cites), log(in_network$total_cites), 2),
     vertex.frame.color = "gray",  # we'll put a frame around it
     vertex.label.color = "black",  # not that ugly blue color for the names
     vertex.label.family = "Helvetica", # not a fan of times in figures
     vertex.label.cex = 0.65,  # a bit smaller too
     vertex.label.dist = 0.5,  # we'll pull the labels a bit away from the vertices
     edge.curved = 0.2, # curved edges is always a nice tough
     edge.arrow.size = 0.1) # arrow size smaller
```

Identify the nodes of RU sociology staff members

```{r}
net1_mat <- igraph::as_adjacency_matrix(net1, sparse = FALSE)

net1_un <- (net1_mat + t(net1_mat)) > 0

net1_un <- graph_from_adjacency_matrix(net1_un, mode = "undirected")

net_sel <- induced.subgraph(net1_un, vids = (in_network$vcol == "#56B4E9"))


plot(net_sel)
```

Now cluster them (based on what though?)

```{r}
clp <- cluster_optimal(net_sel)

plot(clp, net_sel)
```

<br>

## Web scraping

```{r}
# start with clean workspace
rm(list = ls())

# install.packages('data.table')
library(data.table)  # mainly for faster data handling
library(tidyverse)  # I assume you already installed this one!
# install.packages('httr') # we don't need this for now require(httr)
# install.packages("xml2")
require(xml2)
# install.packages("rvest")
require(rvest)
# install.packages("devtools")
require(devtools)
# Note we're doing something different here. We're installing a *latest* version directly from
# GitHub This is because the released version of this packages contains some errors!
# devtools::install_github("jkeirstead/scholar")

require(scholar)

```

```{r}
url <- "https://nos.nl/zoeken?q=jochem+tolsma&page=1"
nos <- read_html(url)
nos

lijst <- nos %>% html_elements("main") %>%
  html_elements("ul") %>%
  html_elements("li")

lijst[1] %>% html_children %>% html_attr(name="href") # nu hebben we de titels van de artikelen

# het kost meestal heel veel tijd om het juiste element te vinden

?rvest

lijst[1] %>% 
  html_children %>%
  html_children %>% 
  html_children %>%
  html_children %>%
  html_children %>%
  html_attr(name="datetime")

# <time> = de naam van het blokje
# datetime = een attribuut van de data in het blokje

lijst[1] %>% html_element("time") 
lijst[1] %>% html_element("time") %>% html_attr(name="datetime")
lijst[1] %>% html_element("time") %>% html_text

artikel1 <- lijst[1] %>% html_children %>% html_attr(name="href")
url <- paste0("https://nos.nl", artikel1)
url

```

## Mijn project - TNSS met 1 wave

### *Testing if RSiena accepts static network data*

```{r}
library(RSiena)

net <- matrix(sample(0:1, size=100, replace=TRUE), nrow=10, ncol=10)
diag(net) <- 0 

net1 <- net2 <- net

beh1 <- sample(1:5, size=10, replace=T)
beh2 <- sample(1:5, size=10, replace=T)

beh <- sienaDependent(as.array(cbind(beh1, beh2)), type= "behavior")


ynet <- array(data = c(net1, net2), dim = c(dim(net1), 2))
ynet <- sienaDependent(ynet)

net <- coDyadCovar(net)

mydata <- sienaDataCreate(beh, ynet)


getEffects(mydata)
```
